---
title: "Assignment4"
author: "Matteo"
date: "2025-01-28"
output: html_document
---

Matteo Ienzi 20270101
Done alone as I was sick and unable to meet with group
https://github.com/MatteoIBiol432/Assignment_4_Solo

```{r}
data <- read.csv("./Cohen_CANCERSEEK_liquid_biopsy_2018_modified.csv")

str(data)
sum(is.na(data))
summary(data)
```

```{r}
boxplot(data[, 5:42])
```
 In Random Forest models, normalization is not required because the algorithm is based on decision trees, which are not sensitive to the scale of the data.

```{r}
dim(data)
table(data$Tumor_type)
```

```{r}
normal_samples <- sum(data$Tumor_type == "Normal")
tumor_samples <- nrow(data) - normal_samples

cat("Number of normal samples:", normal_samples, "\n")
cat("Number of tumor samples:", tumor_samples, "\n")
```

```{r}
train_data <- data[seq(1, nrow(data), by=2), ]
test_data <- data[seq(2, nrow(data), by=2), ]

dim(train_data)
dim(test_data)
```

```{r}
train_data$Tumor_type <- as.factor(train_data$Tumor_type)
train_data_subset <- train_data[, c("Tumor_type", colnames(train_data)[5:42])]
str(train_data_subset)
```

```{r}
library(tree)
model <- tree(Tumor_type ~ ., data = train_data_subset)
summary(model)
par(mar = c(1, 1, 1, 1))
plot(model)
text(model, pretty = 0, cex = 0.5, xpd = TRUE)
```
What protein feature was most influential for classifying samples?
sHER2_aEGFR2_sErbB2

```{r}
test_data$Tumor_type <- as.factor(test_data$Tumor_type)
test_data_subset <- test_data[, c("Tumor_type", colnames(test_data)[5:42])]

predictions <- predict(model, newdata = test_data_subset, type = "class")

confusion_matrix <- table(Predicted = predictions, Actual = test_data$Tumor_type)
confusion_matrix
```

```{r}
misclassification_rate <- sum(predictions != test_data$Tumor_type) / length(predictions)
misclassification_rate
```

```{r}
library(randomForest)
rf_model <- randomForest(Tumor_type ~ ., data = train_data_subset, importance = TRUE)
rf_model
```

```{r}
rf_predictions <- predict(rf_model, newdata = test_data_subset)
confusion_matrix_rf <- table(Predicted = rf_predictions, Actual = test_data$Tumor_type)
confusion_matrix_rf

misclassification_rate_rf <- sum(rf_predictions != test_data$Tumor_type) / length(rf_predictions)
misclassification_rate_rf
```

```{r}
importance_rf <- importance(rf_model)
varImpPlot(rf_model, main = "Feature Importance in Random Forest Model", cex = 0.5, xpd = TRUE)
```

```{r}
train_data_subset$binary <- ifelse(train_data_subset$Tumor_type == "Normal", "Normal", "Cancer")
test_data_subset$binary <- ifelse(test_data_subset$Tumor_type == "Normal", "Normal", "Cancer")

train_data_subset$binary <- as.factor(train_data_subset$binary)
test_data_subset$binary <- as.factor(test_data_subset$binary)
rf_model_binary <- randomForest(binary ~ ., data = train_data_subset, importance = TRUE)
rf_model_binary
```

```{r}
valid_indices <- complete.cases(test_data_subset$binary)
test_data_clean <- test_data_subset[valid_indices, ]

rf_predictions_binary <- predict(rf_model_binary, newdata = test_data_clean)
rf_predictions_binary_clean <- rf_predictions_binary[!is.na(rf_predictions_binary)]
actual_binary_clean <- test_data_clean$binary[!is.na(rf_predictions_binary)]

confusion_matrix_rf_binary <- table(Predicted = rf_predictions_binary_clean, Actual = actual_binary_clean)
print(confusion_matrix_rf_binary)

misclassification_rate_rf_binary <- mean(rf_predictions_binary_clean != actual_binary_clean)
accuracy_rf_binary <- mean(rf_predictions_binary_clean == actual_binary_clean)
cat("Misclassification Rate:", misclassification_rate_rf_binary, "\n")
cat("Accuracy:", accuracy_rf_binary, "\n")

if (length(rf_predictions_binary_clean) == length(actual_binary_clean)) {
    library(irr)
    kappa_score <- kappa2(data.frame(Predicted = rf_predictions_binary_clean, Actual = actual_binary_clean))
    cat("Kappa Score:", kappa_score$value, "\n")
} else {
    cat("Kappa calculation skipped due to mismatched lengths.\n")
}

library(randomForest)
varImpPlot(rf_model_binary, main = "Feature Importance in Random Forest Model", cex = 0.5, xpd = TRUE)

```
Il_8 and Il_6

Interleukin-8 (IL-8): IL-8 is a pro-inflammatory chemokine that plays a critical role in tumor progression by promoting angiogenesis.

Interleukin-6 (IL-6): IL-6 is a cytokine involved in immune response and inflammation but is also known to contribute to cancer development. It promotes tumor proliferation.

This model appears promising for distinguishing cancerous from normal samples based on protein biomarkers in the blood. IL-6 and IL-8 are well-established biomarkers associated with inflammation and tumor progression, making them useful for early cancer detection.